---
title: "Final Markdown Report - Fireflies"
author: "Luke Hatch"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```
# ABSTRACT

# BACKGROUND
Fireflies are among the most mesmerizing insects in the world due to their unique glow in the night sky. Most people would run to avoid insects, but for these they stop and watch. In some cases, they even record what they saw, providing firefly citizen scientist data that supports further research on their behavior. Before they're in the sky, however, they spend the majority of their lives underground developing. Duration of their lives underground is influenced by a variety of factors (Evans et al., 2019, p. 266). Depending primarily on climate and soil conditions, the stages of the firefly life cycles can be altered - taking or adding time (Evans et al., 2019, p. 266). Because these conditions change each year, the timing and abundance of emerged fireflies can also fluctuate. This report uses citizen science data collected in Utah to examine firefly abundance over the past 8 years.

## Question  
How has firefly abundance changed monthly over the last 8 years (2016-2024)?

## Hypothesis 
Due to temperature changes (e.g. global warming), fireflies are expected to emerge sooner in the year as each year passes, as warmer temperatures can accelerate their development and affect the timing of emergence (Evans et al., 2019, p. 266).

## Prediction
If temperature increases have influenced firefly emergence, then peak abundance would be observed earlier in years where the temperature is higher earlier in that year (such as emerging in June vs. July).

# METHODS


## 1st Analysis (Bar Plot)

```{r multiplot}
data <-read.csv("UtahFirefly.csv")


ggplot(data, aes(x= Month))+
  geom_bar(fill = "steelblue3") +
  facet_grid(~ Year) +
  labs(title = "Number of Instances by Month and Year", 
       x = "Month", 
       y = "Number of Instances") +
  theme_minimal()
```

## 2nd Analysis (Scatterplot)

```{r second}

#make plot of Max Julian day
table_Julian <- table(data$Year, data$Julian)
#find the max count each year
max_col <- apply(table_Julian, 1, which.max)
# create a data frame with year and max julian day
Julian <- as.numeric(colnames(table_Julian)[max_col])
Year <- c(2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024)
# Create a color gradient
color_gradient <- colorRampPalette(c("steelblue1", "steelblue4"))
# Generate a vector of colors based on Julian values
cols <- color_gradient(100)[as.numeric(cut(Julian, breaks = 100))]
#plot as scatter plot
plot(Year, Julian, main = "Scatterplot of Peak Counts Each Year",
     col = cols,
     pch = 19,       # filled circles
     cex = 1.5,
     ylab = "Julian Day")      # point size

```

## 3rd Analysis (ANOVA)
```{r, third}
#create data frame 
Julian_dataframe <- data.frame(ColumnA = Julian, ColumnB = Year)
m1 <- lm( ColumnA ~ ColumnB, data=Julian_dataframe)
# run anova test for p value
anova(m1)
summary(m1)
```

# DISCUSSION

## Interpretation of Analysis 1

In analysis 1, results in the bar plots show higher bloodmeal counts in both the + and - WNV areas for the house finch, house sparrow, and american robin. In both the + and - areas, these birds are in the same respective order from highest to lower bloodmeal count. These results from analysis 1 support the hypothesis (house finches are important reservoirs in SLC) on the surface/visual level, but the second analysis clarifies significance into the numerical role of house finches specifically.

## Interpretation of Analysis 2

The second analysis was ran in two parts. The first section of the code chunk was evaluating the data from the house finch for any positive test of WNV at the locations. This produced a statistically significant result (0.0287). After this was determined to be significant, additional analysis was done to determine relativity of each positive test at each location. This was also statistically significant (4.54e-05), meaning there was a higher likelihood of WNV presence where house finches were found. This supports our hypothesis of house finches acting as important reservoirs. One uncertainty, however, could be based on the blood meal collection time period, as there may be different hosts in different seasons or different feeding patterns from the mosquitos.

# CONCLUSION

Overall, the results support the hypothesis - house finches showed significance in acting as reservoirs for WNV. This was shown by the data organized in the bar plot demonstrating higher bloodmeals in both positive and negative areas for WNV in house finches. The second analysis delved deeper into the house finch specifically and tested for numerical significance. It can be concluded that as the number of house finch bloodmeals increases, as does the likelihood of finding WNV. Similarly, the more house finch meals mosquitos have, the more likely the spread - and therefore amplification - of WNV in SLC.

# REFERENCES

1. Komar, N., Langevin, S., Hinten, S., Nemeth, N., Edwards, E., Hettler, D., Davis, B., Bowen, R., & Bunning, M. (2003). Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerging Infectious Diseases, 9(3), 311–322. https://doi.org/10.3201/eid0903.020628

2. ChatGPT. OpenAI, version Nov 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.