---
title: "Warm-Up Mini Report: Mosquitos"
author: "LUKE HATCH"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## BACKGROUND

Fill in some text here that provides background info on the WNV system the blood meal DNA extractions, PCR, sequencing, etc.

```{r viremia, eval = FALSE}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = viremia$Bird,
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# match colors to your species plot
cols_dur <- cols

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols_dur, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

Note: She has examples of the data 

# MAJOR QUESTION

Fill in here the question you want to answer

# METHODS

Fill in here, including overview of procedure and methods in this project
##Fill in name of first ANALYSIS
```{r analysis-1}
# put code here
```
##Fill in name of second ANALYSIS
```{r analysis-2}
# put code here
```
When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
