---
title: "Warm-up Mini-Report: Mosquito Blood Hosts in Salt Lake City, Utah"
author: "Luke Hatch"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

This study investigated the role of different species as West Nile Virus reservoirs in the Salt Lake City area. The proposed question was: Which bird species are the highest amplifying hosts of WNV within Salt Lake City? To determine this, mosquito bloodmeals were collected from various areas known to contain or not contain WNV. These bloodmeals were then amplified through PCR, sequenced, and BLASTed to determine from which organisms they came from. Results showed a positive and significant correlation between WNV presence and house finch bloodmeals. Based on this data, it was determined where house finches are present, there is a higher likelihood of finding WNV. Additionally, the more mosquitos consume blood meals from house finches, the more likely the spread of WNV.

# BACKGROUND

The West Nile Virus (WNV) system is the process in which WNV remains and spreads throughout an environment. Simply put, it is the cycle of the virus that maintains its existence. For example, WNV may be present in a bird, which is then bitten by a mosquito. This transfers the virus to the mosquito, a vector. The mosquito can then feed on another bird or mammal, spreading WNV to a new host. WNV can be fatal in many animals, especially horses (Dr. Greg White). Thus, it is important to understand the movement of this virus through mosquitos and prevent further spread. One way to study this system is through mosquito blood meals. Mosquito blood meals are studied through extracting the DNA from the blood in mosquitos (from hosts they've fed from). This DNA is ran through PCR, where specific sections of the host DNA are amplified for analysis in sequencing. The sequenced DNA can then be matched with known DNA sequences to determine what host species the blood meal came from. This process can be seen in the research of Komar et al. (2003) as they examined American birds with the New York 1999 strain of WNV. Birds tend to have high viremia, meaning they hold higher concentrations of the virus in their blood. This makes them great hosts for WNV and it's spread from mosquito bloodmeals. Komar et al. (2003) examined 25 bird species after infection to see their viremia levels and survival rates of WNV. One species in particular, the House Finch, is a very common sight in Salt Lake City and showed high WNV reservoir competence in their research. Additionally, as seen below house finches tend to "hold onto" the virus for longer. With these findings, we could expect to see higher levels of WNV present in mosquitos' house finch blood meals in our research.

```{r viremia, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# Choose some colors
cols <- c(rainbow(30)[c(10:29,1:5)])  # rainbow colors

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

## Question  
Which bird species are the highest amplifying hosts of WNV within Salt Lake City?

## Hypothesis 
House finches are acting as important reservoirs for WNV in Salt Lake City. 
  
## Prediction
Due to research showing higher viremia in house finches (common species in SLC area), locations known to have WNV will have higher bloodmeal counts from house finches and sparrows than other bird species in the same area. It would also be expected to see higher bloodmeal counts overall in WNV areas compared to non-WNV areas. 

# METHODS

Mosquitos were trapped in various areas known to either have or not have the WNV present. They were trapped using either a Gravid or CO2 trap. Following the collection of mosquitos, the mosquitos were smashed/ground up to extract the DNA from their blood meals. The DNA was amplified and sequenced to compare with known DNA sequences in order to identify what hosts the mosquitos were feeding on. After identifying each species, the data was placed in a file with the DNA sequences and associated species. This data was then analyzed to answer the research question and test for significance. This was done in R-studio (see analysis 1 and 2). 

## 1st Analysis (Barplot)

The code chunk below aggregates the number of host counts per location and produces a bar plot with blood meal count on the x-axis and each species on the y-axis. The barplot shows mosquito blood meals between areas with known WNV-negative pools and areas with known WNV-positive pools among different bird species. The barplot provides a visual comparison to quickly distinguish which bird species had the highest blood meal counts in each area.

```{r first-analysis, warning=FALSE}
## import counts_matrix: data.frame with column 'loc_positives' (0/1) and host columns 'host_*'
counts_matrix <- read.csv("./bloodmeal_plusWNV_for_BIOL3070.csv")

## 1) Identify host columns
host_cols <- grep("^host_", names(counts_matrix), value = TRUE)

if (length(host_cols) == 0) {
  stop("No columns matching '^host_' were found in counts_matrix.")
}

## 2) Ensure loc_positives is present and has both levels 0 and 1 where possible
counts_matrix$loc_positives <- factor(counts_matrix$loc_positives, levels = c(0, 1))

## 3) Aggregate host counts by loc_positives
agg <- stats::aggregate(
  counts_matrix[, host_cols, drop = FALSE],
  by = list(loc_positives = counts_matrix$loc_positives),
  FUN = function(x) sum(as.numeric(x), na.rm = TRUE)
)

## make sure both rows exist; if one is missing, add a zero row
need_levels <- setdiff(levels(counts_matrix$loc_positives), as.character(agg$loc_positives))
if (length(need_levels)) {
  zero_row <- as.list(rep(0, length(host_cols)))
  names(zero_row) <- host_cols
  for (lv in need_levels) {
    agg <- rbind(agg, c(lv, zero_row))
  }
  ## restore proper type
  agg$loc_positives <- factor(agg$loc_positives, levels = c("0","1"))
  ## coerce numeric host cols (they may have become character after rbind)
  for (hc in host_cols) agg[[hc]] <- as.numeric(agg[[hc]])
  agg <- agg[order(agg$loc_positives), , drop = FALSE]
}

## 4) Decide species order (overall abundance, descending)
overall <- colSums(agg[, host_cols, drop = FALSE], na.rm = TRUE)
host_order <- names(sort(overall, decreasing = TRUE))
species_labels <- rev(sub("^host_", "", host_order))  # nicer labels

## 5) Build count vectors for each panel in the SAME order
counts0 <- rev(as.numeric(agg[agg$loc_positives == 0, host_order, drop = TRUE]))
counts1 <- rev(as.numeric(agg[agg$loc_positives == 1, host_order, drop = TRUE]))

## 6) Colors: reuse your existing 'cols' if it exists and is long enough; otherwise generate
if (exists("cols") && length(cols) >= length(host_order)) {
  species_colors <- setNames(cols[seq_along(host_order)], species_labels)
} else {
  species_colors <- setNames(rainbow(length(host_order) + 10)[seq_along(host_order)], species_labels)
}

## 7) Shared x-limit for comparability
xmax <- max(c(counts0, counts1), na.rm = TRUE)
xmax <- if (is.finite(xmax)) xmax else 1
xlim_use <- c(0, xmax * 1.08)

## 8) Plot: two horizontal barplots with identical order and colors
op <- par(mfrow = c(1, 2),
          mar = c(4, 12, 3, 2),  # big left margin for species names
          xaxs = "i")           # a bit tighter axis padding

## Panel A: No WNV detected (loc_positives = 0)
barplot(height = counts0,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (-)",
        xlim = xlim_use)

## Panel B: WNV detected (loc_positives = 1)
barplot(height = counts1,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "Locations WNV (+)",
        xlim = xlim_use)

par(op)

## Keep the colors mapping for reuse elsewhere
host_species_colors <- species_colors
```

## 2nd Analysis (Logistic/Linear Regressions)

This analysis determines whether it is simply the presence or the number of house finch blood meals that can predict whether a site has WNV + pools (binary) or a higher WNV positivity rate (numeric). Essentially, this statistical test allows the formal evaluation of the relationships visualized above in the barplots and determines whether the house finch is influencing whether WNV is or is not found in an area.

```{r}
# second-analysis-or-plot, glm with house finch alone against binary +/_
glm1 <- glm(loc_positives ~ host_House_finch,
            data = counts_matrix,
            family = binomial)
summary(glm1)

#glm with house-finch alone against positivity rate
glm2 <- glm(loc_rate ~ host_House_finch,
            data = counts_matrix)
summary(glm2)
```

# DISCUSSION

## Interpretation of Analysis 1

In analysis 1, results in the barplots show higher bloodmeal counts in both the + and - WNV areas for the house finch, house sparrow, and american robin. In both the + and - areas, these birds are in the same respective order from highest to lower bloodmeal count. These results from analysis 1 support the hypothesis (house finches are important reservoirs in SLC) on the surface/visual level, but the second analysis clarifies significance into the numerical role of house finches specifically.

## Interpretation of Analysis 2

The second analysis was ran in two parts. The first section of the code chunk was evaluating the data from the house finch for any positive test of WNV at the locations. This produced a statistically significant result (0.0287). After this was determined to be significant, additional analysis was done to determine relativity of each positive test at each location. This was also statistically significant (4.54e-05), meaning there was a higher likelihood of WNV presence where house finches were found. This supports our hypothesis of house finches acting as important reservoirs.

# CONCLUSION

Overall, the results support the hypothesis - house finches showed significance in acting as reservoirs for WNV. This was shown by the data organized in the barplot demonstrating higher bloodmeals in both positive and negative areas for WNV in house finches. The second analysis delved deeper into the house finch specifically and tested for numerical significance. It can be concluded that as the number of house finch blood meals increases, as does the likelihood of finding WNV. Similarly, the more house finch meals mosquitos have, the more likely the spread - and therefore amplification - of WNV in SLC.

# REFERENCES

1. Komar, N., Langevin, S., Hinten, S., Nemeth, N., Edwards, E., Hettler, D., Davis, B., Bowen, R., & Bunning, M. (2003). Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerging Infectious Diseases, 9(3), 311–322. https://doi.org/10.3201/eid0903.020628

2. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.